---
title: "A framework for the standard analysis for FFH or research projects"
author: "Suet Li Hooi"
date: "`r Sys.Date()`"
output:
  html_document:
    echo: true
---

Objective: This is a framework on the standard analysis pipeline for any FFH or research projects. Please do note that the codes might still need to be adjusted to suit the structure of the input data in order to run the following analysis successfully. 

Analysis: Alpha diversity (Shannon, Simpson), Beta diversity (PCoA plot based on Aitchison distance), PERMANOVA, Differential abundant (DA) analysis using LEfSe, phylum and genus distribution barplots. 

*Note1: In this example, LEfSe differential abundant analysis is shown, however, many other DA analyses can be used depending on the result and aim of the project (eg. Maaslin2, ANCOMBC, DESeq2, etc.)

*Note2: This analysis can also be used for functional pathway analysis from paprica output, just need to replace the input data with functional pathway input in the following codes.

```{r}
#set your working directory
setwd("~/AMILI_2022/FFH_pipeline")

#load libraries
library(tidyverse)
library(ggplot2)
library(phyloseq)
library(microbiomeMarker)
library(ggsignif)
library(ggpubr)
library(gridExtra)
library(vegan)
library(ggsci)
```


### i. Data cleaning 
```{r}
# load input files - amili 16S v3 processed samples must be obtained from s3://amili-global-archive bucket. Please get the s3 path from the lab team in order to download to RStudio and run the analysis. 

## abundance table - if it's from paprica output, use '*_edge_tally.csv' as the edge_abundance table
##example: 
edge_table <- read.csv("./ffh_input/yoghurt.bacteria.edge_tally.csv", header = T) %>%   select(-X) %>%
  mutate(V1 = gsub(".16S.exp.", "", V1)) %>%
  column_to_rownames("V1") %>%
  replace(is.na(.), 0) %>%
  t() %>%
  as.data.frame() 

## metadata - obtain from the ffh team
##example to upload to RStudio if it's an excel file: 
metadata <- readxl::read_xlsx("./ffh_input/metadata.xlsx", sheet = "metadata") %>%
  column_to_rownames("analysis_ID")

## taxa table - if it's from paprica output, use '*_taxon_map.csv' as the taxa table
##example: 
taxa_table <- read.csv("./ffh_input/yoghurt.bacteria.taxon_map.csv", header = T) %>%
  setNames(.[1,]) %>%
  .[-1,] 

colnames(taxa_table)[1] <- "Index"
colnames(taxa_table)[2] <- "Edge"

rownames(taxa_table) <- NULL

#cleaning up empty rows and renaming taxa levels
taxa_table <- taxa_table %>%
  select(-Index) %>%
  filter(!duplicated(Edge)) %>%
  filter(!is.na(Edge)) %>%
  column_to_rownames("Edge") %>%
  dplyr::rename(Kingdom = superkingdom, Phylum = phylum, Class = class, Order = order, Family = family, Genus = genus, Species = species) %>%
  select(-clade, -strain, -taxon) %>%
  mutate_all(na_if, "") %>%
  replace_na(list(Order = 'Unclassified', Family = 'Unclassified', Genus = 'Unclassified', Species = 'Unclassified'))
```


### ii. Create a phyloseq object (abundance, metadata, taxa) for downstream analysis
```{r}
#create a phyloseq object - phyloseq object needs at least otu_table, taxa_table and sample_data to form
#otu_table is the abundance data of all the samples - contains numerical values

OTU <- otu_table(as.matrix(edge_table), taxa_are_rows = T)

#taxa_table is the table of different taxa classifications - family, genus, species, etc.

TAXA <- tax_table(as.matrix(taxa_table))

#sample_data is the metadata that contains information about the samples such as demographics (age, race, gender, treatment group, etc.)

SAMPLE <- sample_data(metadata)

#after formatting all the input tables for phyloseq, we are now ready to create phyloseq object 
ps <- phyloseq(OTU, TAXA, SAMPLE)

ps

#transform absolute abundance to relative abundance
psr  = transform_sample_counts(ps, function(x) x / sum(x) )
psr

```


### iii. Alpha diversity (Shannon and Simpson)

```{r}
alpha = estimate_richness(ps, measures=c('Shannon', 'Simpson')) 
alpha <- merge(alpha, vag_metadata, by = 'row.names')
```

```{r}
#boxplot for alpha diversities --please refer to ggplot2 R package for the parameters

#plot shannon diversity
Shannon.plot <- 
  alpha %>% 
  ggplot(aes(x=Group, y=Shannon)) + 
  geom_boxplot(aes(color = Group)) + 
  geom_jitter(aes(color = Group), size =0.9) +
  theme_bw() +
  xlab("Group") +
  ylab("Shannon") +
  labs(color='Group') +
  theme(text = element_text(size = 18)) +
  stat_compare_means(aes(group=Group)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle('a) Shannon')
```

```{r}
#plot simpson diversity
Simpson.plot <- 
  alpha %>% 
  ggplot(aes(x=Group, y=Simpson)) + 
  geom_boxplot(aes(color = Group)) + 
  geom_jitter(aes(color = Group), size =0.9) +
  theme_bw() +
  xlab("Group") +
  ylab("Simpson") +
  labs(color='Group') +
  theme(text = element_text(size = 18)) +
  stat_compare_means(aes(group=Group)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle('b) Simpson')

```


```{r, fig.width=12, fig.height=8}
#compile both shannon and simpson diversity into one single plot
alpha.div <- grid.arrange(Shannon.plot, Simpson.plot, ncol = 2) 

print(alpha.div)

ggsave('alpha.div.png', alpha.div, width=12, height=8, units = "in") 
```


### iv. Beta diversity (PCoA plot based on Aitchison distance)

```{r}
#filter rare and low abundance taxa

#remove taxa with abundances below 0.0001 
psr.filt <- phyloseq::genefilter_sample(psr, filterfun_sample(function(x) x >= 0.0001))

psr.filt <- prune_taxa(psr.filt, psr)
psr.filt
```

```{r}
#centred-log ratio (clr) transformation to account for zero-inflated abundance data
psr.filt.clr <- microbiome::transform(psr.filt, 'clr') 

#generate euclidean distance and ordination
euc_dist = phyloseq::distance(otu_table(psr.filt.clr), method="euclidean")

ord_euc = ordinate(psr.filt.clr, method = 'PCoA', distance=euc_dist)
```

```{r, fig.width=6, fig.height=6}
#plot pcoa plot based on aichitson distance
plot_ordination(psr.filt.clr, ord_euc, color = "Group") +  theme(aspect.ratio=1) +
  theme(plot.title = element_text(size = 30)) +
  geom_point(size = 2) +
  labs(color = 'Group') + 
  scale_colour_d3("category20") +
  theme_bw() +
  ggtitle('PCoA plot based on Aitchison distance')
```


### v. Differential abundant analysis using LEfSe

```{r}
#run lefse from microbiomeMarker package using run_lefse function
lefse <- run_lefse(psr.filt,
        wilcoxon_cutoff = 0.05,
        norm = "CPM",
        group = "Group",
        kw_cutoff = 0.05,
        multigrp_strat = TRUE,
        lda_cutoff = 3, 
        taxa_rank = "Species")

lefse 
```

```{r}
#obtaining marker table from the lefse result

marker_lefse <- as(marker_table(lefse), 'matrix') %>%
  as.data.frame() %>%
  mutate(ef_lda = if_else(enrich_group == 'Preterm', -as.numeric(ef_lda), as.numeric(ef_lda)))
```

```{r, fig.width=16, fig.height=12}
#plot barplot for enriched taxa (markers) based on lefse result
marker_lefse %>% 
  ggplot(aes(x = reorder(feature, +ef_lda), y = ef_lda, fill = enrich_group)) +
  geom_bar(stat='identity') + 
  xlab('Taxa') +
  ylab('Effect size (LDA > 3)') +
  coord_flip() + 
  theme_bw() +
  theme(text = element_text(size = 22))

ggsave('marker_lefse.png', width = 16, height = 12)

```


### vi. Phylum and genus distribution barplots 

## a. Phylum
```{r}
#Phylum distribution

# agglomerate taxa
glom <- tax_glom(psr.filt, taxrank = 'Phylum', NArm = FALSE)
psr.melt <- psmelt(glom)

# change to character for easy-adjusted level
psr.melt$Phylum <- as.character(psr.melt$Phylum)

# get the median abundance
psr.melt <- psr.melt %>%
  group_by(Group, Phylum) %>%
  mutate(median=median(Abundance))

#to get the same rows together
psr.melt_sum <- psr.melt %>%
  group_by(Group, Phylum) %>%
  summarise(Abundance=sum(Abundance))
```

```{r}
#plot barplot of phyla abundances 
ggplot(psr.melt_sum, aes(x=Group, y = Abundance, fill = Phylum)) + 
  geom_bar(stat = "identity", aes(fill=Phylum), position = "fill") + 
  labs(x="Sample", y="Abundance") +
  theme_bw() + 
  theme(strip.background = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_d3("category20") +
  ggtitle("Phylum distribution across different samples")
```


## b. Genus
```{r}
#Genus distribution

# agglomerate taxa 
glom.genus <- tax_glom(psr.filt, taxrank = 'Genus', NArm = FALSE)
psr.melt.genus <- psmelt(glom.genus)

# change to character for easy-adjusted level
psr.melt.genus$Genus <- as.character(psr.melt.genus$Genus)
```

```{r}
# get the median abundance
psr.melt.genus <- psr.melt.genus %>%
  group_by(Group, Genus) %>%
  mutate(median=median(Abundance))

# select group mean > 0.25 (depending on your preference)
keep <- unique(psr.melt.genus$Genus[psr.melt.genus$median > 0.25])
psr.melt.genus$Genus[!(psr.melt.genus$Genus %in% keep)] <- "< 0.25"

#to get the same rows together
psr.melt_sum.genus <- psr.melt.genus %>%
  group_by(Group, Genus) %>%
  summarise(Abundance=sum(Abundance))
```

```{r, fig.width=10, fig.height=6}
#plot barplot of genera abundances

ggplot(psr.melt_sum.genus, aes(x=Group, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", aes(fill=Genus), position = "fill") + 
  labs(x="Group", y="Abundance") +
  theme_bw() + 
  theme(text = element_text(size=16)) +
  theme(legend.position = "right", 
        strip.background = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_d3("category20") +
  ggtitle("Genera distribution across different samples")

```


